The error in Bayesian phylogenetic reconstruction when speciation is not instantaneous

### 2018-10-11 article.tex
New version

The PBD model has five parameters, depicted in table \ref{table:parameters}. 
The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
This means that there is a $0.15\thinspace dt$ ($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
probability of speciation completion occurring in an infinitesimal time $dt$. 
We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
We use tree sizes of $n$ of 50, 100 and 200 good taxa.
For each combination of $\lambda$, $\mu$ and $n$, 
we use a speciation initiation rate $b = b_i = b_g$ 
so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
given a $b$, $\lambda$ and $\mu$,
equals the desired number of species $n$.
$b$ is calculated using the PBD R package (\cite{pbd})
for each parameter combination and shown in table \ref{table:pbd_parameters}.
We use $\lambda = 10^9 \approx \infty$ as our control for which the PBD model reduces to the BD model.

@@

We create two data sets: a general one, to explore parameter space, 
and one to investigate the effect of sampling incipient species (see below).
For the general data set, all the trees with the correct number of good species are kept.
There is an additional selection criterion, for the data set to investigate sampling:
to generate that data set, only incipient species trees are kept, 
on which the two sampling methods (see below) result in different species trees.
As sampling will never have an effect for $\lambda = \infty$, 
this parameter value is absent in that data set.

@@

As described above, per alignment, we do four different Bayian analyses,
as we use two different site models and two different clock models.
We know the generative site model (which is JC69) and clock 
model (a strict clock), but want to explore how often the correct model
is indeed preferred. Per alignment, we measure the estimated marginal 
likelihood (which is the probability of the data given the model)
for each of the four models and measure their relative propertions.
To estimate the marginal likelihood, we use the novel 
Nested Sampling approach (\cite{maturana_et_al_2018}),
which we configure to have an effective sample size (in estimating
the marginal likelihood) of at least 200, by using a single-particle
sub-chain with a length equal to the MCMC interval.
\richel{
  There are no rules described in the setup of the NS
  (for example, what should the relation between MCMC chain length
  and NS sub-chain length be?),
  but one can experiment with a sub-chain length and number of particles.
  I will need to do some pilot studies to verify if this is
  the correct way to do and describe this.  
}

@@

For both data sets, we display the nLTT statistics distribution per
biological parameter combination as a violin plot.
\richel{Start new}
We only show the nLTT distributions
that were generated under the (correct) assumptions of a Jukes-Cantor site model
and a strict clock model, separated per sampling method used. 
\richel{End new}
We display the nLTT statistic distributions separated per site or clock model 
in the supplementary information.

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_model_selection.png}
  \caption{
    Model preference on the general parameter set. 
  }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters, see table \ref{table:parameters} \\
    $1000$ & nLTT statistic values \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. Each row will contain one experiment,
    where the columns contain parameters, measurements and diagnostics.
    This table displays the content of the columns. 
    $n$ denotes the number
    of columns a certain item will occupy, resulting in a table of 
    1023 \richel{recalc} columns and 20K rows.
  }
  \label{table:specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\richel{Start new}
The PBD model has five parameters, depicted in table \ref{table:parameters}. 
The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
This means that there is a $0.15\thinspace dt$ ($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
probability of speciation completion occurring in an infinitesimal time $dt$. 
We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
We use tree sizes of $n$ of 50, 100 and 200 good taxa.
For each combination of $\lambda$, $\mu$ and $n$, 
we use a speciation initiation rate $b = b_i = b_g$ 
so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
given a $b$, $\lambda$ and $\mu$,
equals the desired number of species $n$.
$b$ is calculated using the PBD R package (\cite{pbd})
for each parameter combination and shown in table \ref{table:pbd_parameters}.
We use $\lambda = 10^9 \approx \infty$ as our control for which the PBD model reduces to the BD model.

@@

We create two data sets: a general one, to explore parameter space, 
and one to investigate the effect of sampling incipient species (see below).
For the general data set, all the trees with the correct number of good species are kept.
There is an additional selection criterion, for the data set to investigate sampling:
to generate that data set, only incipient species trees are kept, 
on which the two sampling methods (see below) result in different species trees.
As sampling will never have an effect for $\lambda = \infty$, 
this parameter value is absent in that data set.
\richel{End new}

@@

\richel{
  Note there is no twinning, as in the razzo package.
  Already sent an email to discuss this with RSE.
}

\richel{Start new}
As described above, per alignment, we do four different Bayian analyses,
as we use two different site models and two different clock models.
We know the generative site model (which is JC69) and clock 
model (a strict clock), but want to explore how often the correct model
is indeed preferred. Per alignment, we measure the estimated marginal 
likelihood (which is the probability of the data given the model)
for each of the four models and measure their relative propertions.
To estimate the marginal likelihood, we use the novel 
Nested Sampling approach (\cite{maturana_et_al_2018}),
which we configure to have an effective sample size (in estimating
the marginal likelihood) of at least 200, by using a single-particle
sub-chain with a length equal to the MCMC interval.
\richel{
  There are no rules described in the setup of the NS
  (for example, what should the relation between MCMC chain length
  and NS sub-chain length be?),
  but one can experiment with a sub-chain length and number of particles.
  I will need to do some pilot studies to verify if this is
  the correct way to do and describe this.  
}
\richel{End new}

@@

\richel{Start new}
In this article, we showcase the effect of sampling
and the certainty when selecting a (site and clock) model. 
We show the effect sampling has on the inference error,
for the nLTT distributions generated from assuming the (correct) Jukes-Cantor 
site model and a strict clock model. We do so per parameter combination,
displaying the distribution as a violin plot.
In the supplementary information, we display these error distributions 
of the general data set all combined, or separated per site or clock model.
The certainty in the model selection is again shown per parameter setting,
as a stacked bar. 
\richel{
  I am unsure how to depict that stacked bar and its uncertainty.
  Sure, the stacked bars (without uncertainty) will work,
  but information about uncertainty is lost. 
  See figure 'Showing a stacked bar with uncertainty'
  for three equivalent ways to display a stacked bar with uncertainty.
}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_stacked_bar_with_uncertainty.png}
  \caption{
    Showing a stacked bar with uncertainty. These three different ways
    display the same underlying data.
  }
\end{figure}

\richel{End new}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_model_selection.png}
  \caption{
    Model preference on the general data set.
  }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters, see table \ref{table:parameters} \\
    $1000$ & nLTT statistic values \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    $1$ & Marginal likelihood estimate \\
    $1$ & Marginal likelihood estimation uncertainty \\
    $1$ & Marginal likelihood ESS \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. Each row will contain one experiment,
    where the columns contain parameters, measurements and diagnostics.
    This table displays the content of the columns. 
    $n$ denotes the number
    of columns a certain item will occupy, resulting in a table of 
    1023 \richel{recalc} columns and 20K rows.
  }
  \label{table:specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
